---
title: rOpenSci Slack emojis
author: "Lucy D'Agostino McGowan"
date: '2017-07-17'
slug: ropensci-slack-emojis
categories: ["rOpenSci", "rstats", "emojis"]
tags: ["rOpenSci", "rstats", "emojis"]
excerpt: HAPPY world emoji day! In honor of this momentous occasion, I have decided to analyze the emojis used on rOpenSci's Slack.
---

HAPPY world emoji day! `r emo::ji("globe")` `r emo::ji("chicken")` `r emo::ji("calendar")`

In honor of this momentous occasion, I have decided to analyze the emojis used on [rOpenSci](https://ropensci.org)'s Slack. 

```{r, message = FALSE, warning = FALSE}
library("dplyr")
```

`r tufte::margin_note("If you'd like to follow along, go fetch yourself a [Slack token](https://api.slack.com/tokens).")`

```r
token <- "MY_SLACK_API_TOKEN" ## stick your token here
```

We will first use Slack's [`reactions.list`](https://api.slack.com/methods/reactions.list) method.

`r tufte::margin_note("Notice here I am pulling the items from the response and then from each item I am interested in the message reactions. This ignores the reactions on files and comments.")`

```r
req_lst <- httr::POST(
  "https://slack.com/api/reactions.list",
  body = list(
   token = token,
   count = 500
   )) %>%
  httr::content() %>%
  .$items %>%
  purrr::map(~.[["message"]][["reactions"]])
  
```

Let's pull out the `name` and `count` of each emoji used and stick it in a `tibble` `r emo::ji("tada")`. 
```r
tbl <- tibble::tibble(
  name = purrr::map_chr(purrr::flatten(req_lst), "name"),
  count = purrr::map_int(purrr::flatten(req_lst), "count")
)
```

One of the most delightful features in Slack is the ability to create custom emojis! In order to be able to display both ordinary and custom emojis, I can pull in a list of all emojis we have customized in the rOpenSci Slack team using the [`emoji.list`](https://api.slack.com/methods/emoji.list) method.

<span class = "marginnote">
Notice here I am using the `tibble::enframe()` function. This is an awesome way to convert a vector or list to a two-column data frame.
</span>
```r
emojis_tbl <- httr::POST(
  "https://slack.com/api/emoji.list",
  body = list(token = token)
  ) %>%
  httr::content() %>%
  .$emoji %>%
  tibble::enframe() %>%
  mutate(value = unlist(value))
```

```{r, echo = FALSE}
load("../../static/data/ropensci_emojis.rda")
load("../../static/data/ropensci_tbl.rda")
```

The `emoji_tbl` data frame contains the name of each custom emoji and a link to their associated image. Here I create a small function that will either read that image using the [`magick`](https://github.com/ropensci/magick) package or, if it is an ordinary emoji, use the [`emo`](https://github.com/hadley/emo) package to look it up.

```{r}
read_emoji <- function(x, y) {
  if (!is.na(x)) {
    y <- which(x == x)
    magick::image_read(x)
  } else {
    emo::ji(y)
  }
}
```

Let's do a wee bit of data wrangling `r emo::ji("tractor")` to sort out which emojis are used the most.

```{r}
emoji_tbl <- tbl %>%
  group_by(name) %>%
  mutate(count = sum(count)) %>%
  distinct() %>%
  arrange(desc(count)) %>%
  head(10) %>%
  left_join(emojis_tbl, by = "name") %>%
  mutate(emoji = purrr::map2(value, name, read_emoji))
```

I have written another small function to make sure the custom emojis print properly when I render my output.

`r tufte::margin_note(paste("Note, if you are doing this for your blog, rather than saving a temporary file as I have demonstrated here, you should save this as a relative file path in your blog", emo::ji('deciduous_tree'), "."))`

```{r, eval = FALSE}
render_emoji <- function(x, y) {
  if (inherits(x, "magick-image")) {
    tmp <- tempfile(fileext = paste0(".gif"))
    x <- magick::image_scale(x, "20x20")
    magick::image_write(x, path = tmp)
    emoji <- rep(glue::glue("![]({tmp})"), as.integer(y/7))
    print(glue::glue("{glue::collapse(emoji)}: {y}\n\n"))
  } else {
    print(glue::glue(
      "{glue::collapse(rep(x, as.integer(y/7)))}: {y}\n\n"
      ))
  }
}
```

```{r, echo = FALSE}
render_emoji <- function(x, y) {
  if (inherits(x, "magick-image")) {
    tmp <- paste0("../../static/images/ropensci_custom_emoji", y, ".gif")
    x <- magick::image_scale(x, "25x25")
    magick::image_write(x, path = tmp)
    tmp <- paste0("/images/ropensci_custom_emoji", y, ".gif")
    emoji <- rep(glue::glue("![]({tmp})"), as.integer(y/7))
    print(glue::glue("{glue::collapse(emoji)}: {y}\n\n"))
  } else {
    print(glue::glue("{glue::collapse(rep(x, as.integer(y/7)))}: {y}\n\n"))
  }
}
```

Now let's walk it out `r emo::ji("dancer")`.

```{r, results = "asis"}
purrr::walk2(emoji_tbl$emoji, emoji_tbl$count, render_emoji)
```

How delightful! It is no surprise that the `r emo::ji("wave")` is the most popular emoji used, since rOpenSci is an extraordinarily welcoming crew! In fact the community organizer, [Stefanie](https://twitter.com/StefanieButland), just wrote a blog post about our Slack [#welcome channel](https://blog.trelliscience.com/the-value-of-welcome/). I am (obviously) very proud to see that `r emo::ji("rooster")` makes a strong appearance ([#rchickenladies](https://twitter.com/search?src=typd&q=%23rchickenladies)), and of course delighted to see our Aussie friends getting represented by the Aussie party parrot. 

