---
author: "Lucy D'Agostino McGowan"
comments: true
date: 2017-03-04T19:23:45-06:00
draft: false
image: ""
menu: ""
share: true
tags:
- xkcd
- thankyou
title: "xkcd thank you note"
excerpt: "Exploring the relationship between my time spent on statistics & my time spent on xkcd."
---

I've been looking for a way to incoporate xkcd in my #thankyou series, and then I came across this gem:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Did you know that there is a package in R to make your graphs in the style of xkcd comics? <a href="https://twitter.com/hashtag/Rstats?src=hash">#Rstats</a> <a href="https://t.co/29Cnp9t20N">https://t.co/29Cnp9t20N</a> <a href="https://t.co/iY1UsFdEfJ">pic.twitter.com/iY1UsFdEfJ</a></p>&mdash; Maria Gatta (@M_Gatta) <a href="https://twitter.com/M_Gatta/status/832952195784044545">February 18, 2017</a></blockquote> <script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

What could be better? [xkcd](http://xkcd.com) has been there during some of my most [*finest*](http://livefreeordichotomize.com/2016/12/15/hill-for-the-data-scientist-an-xkcd-story/) statistical moments. I decided to use the [xkcd](https://cran.r-project.org/web/packages/xkcd/xkcd.pdf) package to explore the relationship between the time I've spent on statistics and the time I've spent on xkcd & express my extreme graditude for xkcd.

```{r, message = FALSE, warning = FALSE}
library('ggplot2')
library('xkcd')
load("../../data/xkcd_data.rda")
```


```{r, message = FALSE, warning = FALSE}
set.seed(916)
xkcdlabel <-  function(text,x,y, above= FALSE) {
  if (above == TRUE){
    add = 4
    beg = 3
  } else {
    add = -4
    beg = -3
  }
  te1 <- annotate("text", x=x, y = y , label=text,family ="xkcd")
  list(te1,
       xkcdline(aes(xbegin=xbegin, ybegin= ybegin, xend=xend,yend=yend),
                data.frame(xbegin=x,ybegin= y+beg , xend=x,yend=y+add), xjitteramount = 0.5))
}

p <- ggplot(data) + 
  geom_smooth(mapping=aes(x=time, y =spent, group = stat, linetype = stat), 
              se = FALSE,
              method="loess",
              span = 0.25) +
  xkcdaxis(c(2008,2018),c(0,44)) +
  xkcdlabel("BU\n SIBS",x=2011.5,y = 12) +
  xkcdlabel("1st year comps\n closed book",x=2015.3,y = 1.5, above = TRUE) +
  xkcdlabel("2nd year comps\n open book",x=2016.7,y = 42) +
  xkcdlabel("Start \ncollege",x = 2008.6,y = 8) +
  scale_linetype_discrete(name = "Time spent on") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = c(.25,.75)) +
  ggtitle("Thank you, xkcd!")
p



```

